Metabolites reflect variability introduced by mesoscale eddies in the North Pacific Subtropical Gyre

### 2024-10-29 manuscript/response_to_reviewers.txt

> Software: Give the version number of the software used (e.g. Proteowizard, msConvert, 
> XCMS...)
Done. Software versions for Proteowizard, Skyline, MSDIAL, R, XCMS, and vegan have been
added to the manuscript.

> Statistics: It would facilitate better interpretation if the statistical method used 
> for comparisons would be mentioned in the respective sections of results.
We have tried to add statistical methods used for comparisons in each section of the
results where applicable in this revision, largely annotating the places in which 
R-squared and p-values were obtained from PERMANOVAs. We have also added the sample
size to the Statistics section of the Results.

> Line 239 Typo r = 0.697
We have switched to using a 0 in front of Pearson's correlation coefficient throughout
the manuscript to follow modern conventions.

> Line 266: From the figures in the supporting material, it does not become evident that 
> Particulate caron and metabolite concentrations are connected. There could be nested 
> effects due to the selection of data points.
It is unclear what the reviewer means by the selection of data points. We are aware that
the strong trend between particulate carbon and metabolite concentration is largely due
to the effect of depth and that was our inspiration for performing both the overall
regression and independent regressions at each depth.

### 2024-10-29 manuscript/manuscript.Rmd
Copyedits for readability

Samples were also processed via the untargeted workflow detailed in @Kumler2023. In short, XCMS (version 3.10.0, @R-xcms) was used for feature detection with CentWave peakpicking, Obiwarp alignment, and peak density correspondence/grouping. This was followed by manual inspection of the extracted ion chromatogram for each mass feature by an expert for quality with only those peaks classified as "Good" used in the analysis here. Features were then matched with and normalized to an internal standard (Supplemental Table 1) as described above. Scripts used for this analysis are available at https://github.com/wkumler/MesoscopeMetabolomicsManuscript/tree/master/untargeted.

@@

In all of the datasets analyzed here, we detected a significant difference in the composition of the metabolome between the adjacent eddies. This effect was strongest in the samples taken during the Lagrangian stations at the center of each eddy in the 2017 MESO-SCOPE cruise, with nearly a quarter of the total variance explained by the eddy from which the samples were taken. In the samples taken along the eddy transect, the largest effect was detected in the deepest samples taken from 175 meters and from the DCM, with much less of a response at 25 meters. In the HEE samples, however, the opposite was true with a surprisingly larger SLA response at 25 meters than that at the DCM. In each case, the differences introduced by SLA were smaller than those of depth when compared directly, with even the high-resolution sampling around the DCM finding slightly more variance explained by the 40 meter difference in sampling depth than the 40 centimeter difference in SLA. This result agrees well with prior research by @Heal2021 and @Kumler2023 showing the large effect of sampling depth on the metabolome. Time of day was included as a significant factor in the MESO-SCOPE transect, with trehalose and sucrose in particular showing large differences between 6AM and 6PM as has been noted previously [@Muratore2022], though this effect was diluted at 175 meters.

Although possible that the observed differences in metabolomes across adjacent eddies differing in polarity are due to latitudinal shifts or simply background variation in the gyre environment, the transect data in particular implies that this is unlikely to be the case. Samples taken when the absolute value of the SLA was less than 5 cm were more similar to each other than to those samples taken within the eddies despite larger differences in latitude. The transect sampled both outside of each eddy and between the two across a large spatial gradient (~4° latitude and 2° longitude), yet the non-eddy (absolute value of SLA < 5 cm) samples from all three locations grouped together, showed similar median metabolite concentrations, and were visually similar in the most abundant metabolite composition. These results imply that strong cyclonic and anticyclonic eddies represent endpoints in NPSG composition, in agreement with results from @Barone2019 which found the most extreme values in the Hawaii Ocean Time series data typically detected when eddies passed over Station ALOHA. We also detected the largest differences in the DCM metabolome at the exact center of the eddy, with Station 12 highly distinct from even the nearest stations, while no such stark difference was found in the anticyclone (Figures 1 and 2). This perhaps indicates that the exact center of the cyclone has a unique metabolome at the DCM relative to the rest of the transect, rather than existing as a smooth continuation.

@@

Given the confluence of both depth and SLA signals and the way that the median metabolite plots of Figure 2 would confound the SLA signal with the depth variance, we used k-means clustering to group similar compounds and provide a reduced dimensionality space for visualization. This revealed four major patterns of metabolite response, with a majority of compounds responding to eddy state (Clusters 1, 2, and 4 in Figure 5). It is interesting to note that the sole cluster in which abundances in the anticyclone are greater than those of the cyclone is also the only cluster to generally increase in abundance with depth (Cluster 4), perhaps indicating that compounds more concentrated in the anticyclone are the same kind of degradation products and recalcitrant carbon typically found at depth.

>>>>

Samples were also processed via the untargeted workflow detailed in @Kumler2023. In short, XCMS (version 3.10.0, @R-xcms) was used for feature detection with CentWave peakpicking, Obiwarp alignment, and peak density correspondence/grouping. This was followed by manual inspection of the extracted ion chromatogram for each mass feature by an expert for quality with only those peaks classified as "Good" used in the analysis here. Features were then matched with and normalized to an internal standard (Supplemental Table 4) as described above. Scripts used for this analysis are available at https://github.com/wkumler/MesoscopeMetabolomicsManuscript/tree/master/untargeted.

@@

In all of the datasets analyzed here, we detected a significant difference in the composition of the metabolome between the adjacent eddies. This effect was strongest in the samples taken during the Lagrangian stations at the center of each eddy in the 2017 MESO-SCOPE cruise, with nearly a quarter of the total variance explained by the eddy from which the samples were taken. In the samples taken along the eddy transect, the largest effect was detected in the deepest samples taken from 175 meters and from the DCM, with much less of a response at 25 meters. This lack of SLA effect at the surface is unsurprising given the similarity in the biogeochemistry of the well-lit upper ocean previously reported [@Barone2022]. In the HEE samples, however, the opposite was true with a surprisingly larger SLA response at 25 meters than that at the DCM. In each case, the differences introduced by SLA were smaller than those of depth when compared directly, with even the high-resolution sampling around the DCM finding slightly more variance explained by the 40 meter difference in sampling depth than the 40 centimeter difference in SLA. This result agrees well with prior research showing the large effect of sampling depth on the metabolome [@Heal2021; @Kumler2023; @Bent2024]. Time of day was included as a significant factor in the MESO-SCOPE transect, with trehalose and sucrose in particular showing large differences between 6AM and 6PM as has been previously noted [@Muratore2022], though this effect was diminished at 175 meters.

Although possible that the observed differences in metabolomes across adjacent eddies differing in polarity are due to latitudinal shifts or simply background variation in the gyre environment, the transect data in particular implies that this is unlikely to be the case. Samples taken when the absolute value of the SLA was less than 5 cm were more similar to each other than to those samples taken within the eddies despite larger differences in latitude. The transect sampled both outside of each eddy and between the two across a large spatial gradient (~4° latitude and 2° longitude), yet the non-eddy (absolute value of SLA < 5 cm) samples from all three locations grouped together, showed similar median metabolite concentrations, and were visually similar in the most abundant metabolite composition. These results imply that strong cyclonic and anticyclonic eddies represent endpoints in NPSG composition, in agreement with results from @Barone2019 which found the most extreme values in the Hawaii Ocean Time series data typically detected when eddies passed over Station ALOHA. We also detected the largest differences in the DCM metabolome at the exact center of the eddy, with Station 12 highly distinct from even the nearest stations, while no such stark difference was found in the anticyclone (Figures 1 and 2). This indicates that the exact center of the cyclone has a unique metabolome at the DCM relative to the rest of the transect, rather than existing as a smooth gradient with SLA.

@@

Given the confluence of both depth and SLA signals and the way that the median metabolite plots of Figure 2 would confound the SLA signal with the depth variance, we used k-means clustering to group similar compounds and provide a reduced dimensionality space for visualization. This revealed four major patterns of metabolite response, with a majority of compounds responding to eddy state (Clusters 1, 2, and 4 in Figure 5). It is interesting to note that the sole cluster in which abundances in the anticyclone are greater than those of the cyclone is also the only cluster to generally increase in abundance with depth (Cluster 4), likely because compounds more concentrated in the anticyclone are the same kind of degradation products and recalcitrant carbon typically found at depth.
